---
title: "Data visualisation with ggplot"
author: "Ilaria Torre"
date: "21 February 2019"
output:
  ioslides_presentation:
    fig_height: 2
    fig_width: 4
    smaller: yes
    widescreen: yes
  beamer_presentation: default
fontsize: 10pt
---

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(echo = TRUE, width = 40, breaklines = T, tidy.opts = list(width.cutoff = 40), message=FALSE, warning = FALSE, tidy = TRUE, fig.align = "center")

```

## Sources

- ggplot documentation: <http://docs.ggplot2.org/current/>
- easier tutorials: <http://www.noamross.net/blog/2012/10/5/ggplot-introduction.html>, 
- R Markdown (the tool I used to make these slides), also an R package
- grammar of graphics: <https://www.springer.com/gp/book/9780387245447>


## Installation

```{r}
# install.packages("ggplot2")
library(ggplot2)
# View(iris)
# ggthemr_reset()
```

- Pros: part of tidyverse, consistent grammar, makes sense, powerful, active community
- Cons: more difficult to start with, computationally intensive 

## Layers

Plots are built as a series of layers, e.g.:

- One layer for the box that will contain the plot
- One layer for points
- One layer for lines
- One layer for axes
- One layer for annotations
...

Great introduction: <http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html>

## Iris data set

- 150 rows
- 4 measurements
- 1 factor (species) with 3 levels

```{r}
View(iris)
```


## Scatterplot

```{r}
ggplot()
```

Nothing is happening: ggplot is just the first layer

## Scatterplot

```{r}
ggplot() + 
  geom_point(data = iris, aes(x = Sepal.Length, y = Sepal.Width))
```

## Scatterplot

Plots can be assigned to variables too:

```{r}
g <- ggplot()
```

----

And you can add new layers at any point:

```{r}
g + 
  geom_point(data = iris, aes(x = Sepal.Length, y = Sepal.Width))
```

## Scatterplot

Adding trendline

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth()
```

## Scatterplot

Regression line

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth(method = lm)
```

## Scatterplot

Get rid of the error

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F)
```

## Plot using stat function

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  stat_summary(geom = "point", fun.y = mean)
```

Be careful where you put data and aes: only if all layers have the same data, x and y, then you can put these inside ggplot()

## Plot using stat function

fun.y = mean is calling mean(.) on the y-axis.

You can replace mean() with e.g. median() or even your own function

## Scatterplot

Adding the grouping factor with colour

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  stat_summary(geom = "point", fun.y = mean)
```

## Scatterplot

Adding the grouping factor with shape

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, shape = Species)) + 
  stat_summary(geom = "point", fun.y = mean)
```

## Scatterplot

Colour AND shape, why not!

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, shape = Species, color =  Species)) +
  stat_summary(geom = "point", fun.y = mean)
```

## Barplot

``` {r}
ggplot(data = iris, aes(x = Species, y = Petal.Width)) + 
    stat_summary(geom = "bar", fun.y = mean)

 # geom_bar(stat = "identity")
  
```

## Barplot

```{r}
ggplot(data = iris, aes(x = Species, y = Petal.Width, fill = Species)) + 
  stat_summary(geom = "bar", fun.y = mean)
```

Notice: "fill" instead of "color"


## Error bars

```{r}
iris_error <- data.frame(Species = c("setosa", "versicolor", "virginica"), Petal.Width = c(0.246, 1.326, 2.026))
View(iris_error)

g <- ggplot(data = iris_error, aes(x = Species, y = Petal.Width, fill = Species)) + 
  geom_bar(stat = "identity")
g
```

## Error bars

- With a similar data frame + standard error, we can add error bars
- Function "summarySE" can do it for us 
- Source: http://www.cookbook-r.com/Manipulating_data/Summarizing_data/ 

```{r}
source('summarySE.R') 

iris_error <- summarySE(iris, measurevar = "Petal.Width", groupvars = "Species")

```

## Error bars

```{r}
g <- ggplot(data = iris_error, aes(x = Species, y = Petal.Width, fill = Species)) + 
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = Petal.Width - se, ymax = Petal.Width + se), width = .4)
g
```

## Error bars

Instead of using the summarySE function, you can create a summary data frame with tidyverse. It has the advantage of using the same grammar as ggplot, and is faster.

```{r}
# install.packages("tidyverse") 
library(tidyverse)

iris2 <- iris %>% 
	group_by(Species) %>% 
	summarise(
	  N = n(),
		mean = mean(Petal.Width), 
		sd = sd(Petal.Width),
		se = sd / sqrt(50),
		ci = qt(.975, N - 1) * se ) 
```


## Boxplot

```{r}
ggplot(data = iris, aes(x = Species, y = Petal.Width, fill = Species)) + 
  stat_boxplot() 
```

## Line plot

```{r}
ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line")

```

What happens if you write `+ geom_line()' instead of 'stat_summary()?

## Line plot

Lines and points

```{r}
ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line") + 
  stat_summary(fun.y = mean, geom = "point", aes(shape = Species))
```

## Messing around with parameters

Base plot:

```{r}
g <- ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line") + 
  stat_summary(fun.y = mean, geom = "point", aes(shape = Species))
g
```


## Messing around with parameters

- Changing the dimensions of the axes

```{r}
g + 
  xlim(0, 10) + 
  ylim(0,10)
```

## Messing around with parameters

- Changing the size of the elements in the graph

```{r}
ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length,color = Species)) + 
  stat_summary(fun.y = mean, geom = "line", size = 2) + 
  stat_summary(fun.y = mean, geom = "point", aes(shape = Species), size = 3)
```

## Messing around with parameters

- Changing the size of the text around the graph (theme layer)

```{r}
g + theme(axis.text = element_text(size = 14), 
legend.text = element_text(size = 14), 
legend.title = element_text(size = 14), 
axis.title = element_text(size = 14))

```

## Messing around with parameters

- Changing the axes names

```{r}
g + 
  labs(x = "Sepal Width", y = "Sepal Length")
```

## Messing around with parameters

- Adding a title

```{r}
g + 
  labs(x = "Sepal Width", y = "Sepal Length", title = "Iris plot")
```

## Messing around with parameters

- Adding a title in a different way

```{r}
g + 
  ggtitle("Iris plot 2") 
```


## Messing around with parameters

- Custom colours

```{r}
coll <- c("violetred", "darkblue", "lightgreen")
```

Allowed names: <http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf>

## Messing around with parameters

- Custom colours

```{r}
g + 
  scale_color_manual(values = coll)
```

## Messing around with parameters

- Semi Custom colours

```{r}
g + 
  scale_color_brewer(palette = "Spectral")
```

Some palette choices here: <http://www.r-bloggers.com/choosing-colour-palettes-part-ii-educated-choices/>

## Messing around with parameters

- Grey scales

```{r}
g + 
  scale_color_grey()
```

## Messing around with parameters

- Change legend title and names

```{r}
g <- ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line") + 
  stat_summary(fun.y = mean, geom = "point", aes(shape = Species)) + 
  labs(x = "Sepal Width", y = "Sepal Length") + 
  ggtitle("Iris plot 2")

g
```

## Messing around with parameters

- Change legend title and names

```{r}
g + 
  scale_color_grey(name = "New Legend Title", labels = c("se", "ver", "vir"))
```

But this creates 2 different legends!

## Messing around with parameters

- The reason is that we only specified the "coloured" layer to have a new legend
- We need to specify the "shaped" layer as well!

```{r}
g + 
  scale_color_grey(name = "New Legend Title", labels = c("se", "ver", "vir")) +
  scale_shape_discrete(name = "New Legend Title", labels = c("se", "ver", "vir"))
```

## Messing around with parameters

Or define the labels in a variable

```{r}
new_labels <- c("se", "ver", "vir")

g + 
  scale_color_grey(name = "New Legend Title", labels = new_labels) +
  scale_shape_discrete(name = "New Legend Title", labels = c("se", "ver", "vir"))
```

---

```{r}
g <- g + 
  scale_color_grey()

g
```

## Messing around with parameters

- Print-friendly

```{r}
g + 
  theme_bw()

```

## Messing around with parameters

- Delete legend or part of

```{r}
g + 
  guides(color = "none", shape = "none") +
  theme_bw()
```

## Messing around with parameters

- Delete legend or part of:
  Just leave the "" empty



## Additional themes

- Package 'ggthemr': <https://github.com/cttobin/ggthemr>

```{r}
# This is not on CRAN yet. You need to install via
# devtools::install_github('cttobin/ggthemr')
library(ggthemr)
ggthemr('fresh')
g
ggthemr_reset()

```

## Additional themes

- Collection of themes, including a replica of Stata graph schemes: <https://github.com/jrnold/ggthemes>
- APA-themed ggplot template: <https://sakaluk.wordpress.com/2015/04/13/2-the-ggplot2-package-your-gateway-drug-to-becoming-an-r-user/>

## Additional themes

Example of the aforementioned APA theme:

```{r}
# windowsFonts(Times=windowsFont("TT Times New Roman")) 

apatheme = theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(),
        text = element_text(family="Times"),
        legend.title = element_blank())

g + 
  apatheme
```

Note: the font family argument caused the following error in my machine: "Font family not found in Windows font database".
I solved it by running the windowsFonts() function just above. You might not encounter this error though!

## Additional themes

Game of thrones palette!! <https://github.com/aljrico/gameofthrones>

```{r}
# install.packages("gameofthrones")
library(gameofthrones)

t <- ggplot(diamonds, aes(factor(color), fill=factor(cut))) +  
  geom_bar(colour = "black") +
  theme_minimal()

t

```

## Additional themes

```{r}

t + 
  scale_fill_got(discrete = TRUE, option = "Daenerys")
```


## Additional themes

Harry Potter palette!!! <https://github.com/aljrico/harrypotter>

```{r}
# install.packages("harrypotter")
library(harrypotter)

t + 
  scale_fill_hp(discrete = TRUE, house = "slytherin")
```
 


## Faceting

```{r}
g + 
  facet_wrap(~Species)
```

## Faceting

- 2 grouping variables
- Iris only has one so I will add another one

```{r}
iris2 <- iris
iris2$fac <- c("a", "b", "c")

View(iris2)
```



## Faceting

```{r}
g <- ggplot(data = iris2, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line") + 
  stat_summary(fun.y = mean, geom = "point", aes(shape = Species)) + 
  theme(axis.text = element_text(size = 14), legend.text = element_text(size = 14), legend.title = element_text(size = 14), axis.title = element_text(size = 14)) + 
  labs(x = "Sepal Width", y = "Sepal Length") + 
  ggtitle("Iris plot 2") + 
  scale_color_grey() + 
  theme_bw()
```

```{r}
g + 
  facet_grid(Species~fac)

```


## Saving plot 


```{r}

ggsave(g, file = "g.pdf")
ggsave(g, file = "g.png")

```

You can also run the regular pdf() or png() functions. ggsave will understand the format you want to save in from the file name extension.

## Saving plot 


```{r}

ggsave(g, file = "g.png", width = 10, height = 4, units = "cm")

```

## Questions?

Thank you!

Dr. Ilaria Torre
torrei@tcd.ie

<https://edge-research.eu/fellows/ilaria-torre/>
